\acknowledgments
My four years of Ph.D. life in Amsterdam is difficult but productive. In the beginning, I didn't have much confidence in myself for the transition from an experimentalist to a data analyst (statistician). However, during this process, I learned a lot on statistics, and have developed several statistical methods for the data fusion of heterogeneous data sets. These cannot be done without the help and support of a lot of people. Here I would like to express my gratitude to you. 

First, I would like to say thanks to Age Smilde and Johan Westerhuis for your four years' supervision. I appreciate that you accepted me to do my Ph.D. research in the BDA group. And thanks so much for all the discussions we have during the last four years and your insightful comments on my work. These helped me a lot in learning statistics and doing research.

I would also like to say thanks to Huub Hoefsloot and Age Smilde for the matrix algebra course. Matrix algebra is the foundation of statistics and numerical computation. A solid understanding of this subject helps me a lot on the algorithm development for my research. Also, I would like to acknowledge Gooitzen Zwanenburg for your help in solving my problems on the computer, software, Internet, the Dutch translation of the thesis summary, and many other things. Furthermore, I want to thank Dicle, Chloie, Sandra, and Maryam. I am glad to be in the same office as you. Thanks for the discussion we have about life, research and many other topics. Also, thanks for all the members in the BDA group for the coffee break conversations, the presentations, and discussions during the group meetings.

Part of my Ph.D. research has collaborated with Lodewyk Wessels, Nanne Aben from the Netherlands Cancer Institute (NKI) and Patrick J.F. Groenen from the Erasmus University. Thanks to Lodewyk Wessels and Nanne Aben for your real biological data sets, your informative comments on our paper and all the biological interpretations. Also, thanks to Patrick for your enlightening comments on the algorithm section of our paper and for your work on the GDP penalty, which is used throughout my thesis.

I would also like to acknowledge the China Scholarship Council for financial support. The scholarship makes it possible for me to study in Amsterdam.

Lastly, I would like to thank my family and my girlfriend for your love and encouragement.

\vskip 4em
\hfill Yipeng Song
\vskip 1em
\hfill May 21, 2019
